---
title: "Fire hotspots in Edmonton, 2024"
author: "Matt Ashby"
date: "March 2025"
format:
  html:
    embed-resources: true
    fig-cap-location: top
execute:
  echo: false
---


::: {.callout-important}
#### This is an example report used for teaching purposes

This is an example report used to help students understand how different types of report can be created using the Quarto markup language and the R programming language. **This contents of this report should not be relied upon for operational use.**
:::


```{r setup}
#| include: false

pacman::p_load(ggspatial, gt, patchwork, scales, sf, sfhotspot, tidyverse)
```


```{r load-data}
#| include: false
#| cache: true

# CO-ORDINATE REFERENCE SYSTEM
# All spatial datasets will be transformed to use EPSG:3780 which is appropriate
# for Alberta and uses metres as its unit of measurement

## INCIDENTS -------------------------------------------------------------------

# DOWNLOADING THIS DATA
# Edmonton Fire Rescue data is hidden behind a web form, so cannot easily be 
# downloaded directly from a URL (although there are ways round this with the
# httr2 and rvest packages, those are complicated). The dataset is also very 
# large (about 300MB). Since it must be downloaded manually, we are going to
# download it in a web browser and then store it in the data/raw directory of the workspace.

# MISSING VALUES
# A small proportion of rows in the data have missing co-ordinates. We will need
# to remove these before analysis, but it would be useful to point out how many
# rows have been removed. To enable this we will store the data including the
# rows with missing co-ordinates, then create a new object that does not include
# those rows and will be used for most of the analysis.

# Source: https://data.edmonton.ca/d/7hsn-idqi
incidents <- here::here("data", "raw", "Fire_Response__Current_and_Historical_20250309.csv") |> 
  read_csv() |> 
  # Remove duplicate columns (these are usually two columns storing the same
  # data, but in different formats)
  select(
    -dispatch_month, 
    -dispatch_day, 
    -dispatch_month_name, 
    -dispatch_dayofweek, 
    -dispatch_date, 
    -dispatch_datetime, 
    -event_close_date, 
    -event_close_datetime, 
    -geometry_point
  ) |> 
  filter(
    # Remove 'incidents' that are actually fire crews telling the control room 
    # that they are engaged in training (which means they might not be available
    # to respond to incidents).
    !event_description %in% c("TRAINING/MAINTENANCE", "MESS"),
    # Remove 'incidents' that are actually fire crews notifying the control room
    # that they are, for example, attending a school to deliver some fire-safety
    # training
    !event_type_group %in% c("CM", "DR", "86", "FP", "T0", "MP"),
    # Remove incidents that did not occur in 2024
    dispatch_year == 2024
  )

fires <- incidents |> 
  filter(event_type_group %in% c("FR", "OF", "VF")) |> 
  # Categorise different types of fire
  mutate(
    # Extract the first two digits of `event_type_code`, which indicates the
    # type of fire involved in an incident. All structure-fire codes start "69",
    # so we further separate those into residential and other structure fires.
    code2 = str_sub(event_type_code, end = 2),
    # Set the fire types
    fire_subtype = case_when(
      code2 == "57" ~ "explosion",
      code2 == "67" ~ "outside fire",
      code2 == "71" ~ "vehicle fire",
      code2 == "82" ~ "vegetation/woodland fire",
      code2 == "69" & str_detect(event_type_description, "RESIDENTIAL") ~ 
        "residential structure fire",
      code2 == "69" ~ "other structure fire",
      .default = "other non-structure fire"
    )
  ) |> 
  # Remove rows with missing co-ordinates
  drop_na(longitude, latitude) |> 
  # Convert to SF object
  st_as_sf(coords = c("longitude", "latitude"), crs = "EPSG:4326") |> 
  # Transform data to use local co-ordinate reference system
  st_transform("EPSG:3780")



## EVERYTHING ELSE -------------------------------------------------------------

# Source: 
fire_stations <- read_sf("https://data.edmonton.ca/resource/b4y7-zhnz.geojson") |> 
  select(name, address) |> 
  # Transform data to use local co-ordinate reference system
  st_transform("EPSG:3780")

# Source: 
neighbourhoods <- read_sf("https://data.edmonton.ca/resource/65fr-66s6.geojson") |> 
  # Transform data to use local co-ordinate reference system
  st_transform("EPSG:3780")

# Create a single polygon showing the boundary of the city
edmonton <- st_union(neighbourhoods)

# Create polygons for each district in the city
districts <- summarise(
  neighbourhoods, 
  across(geometry, st_union), 
  .by = district
)
```


This report has been prepared for Edmonton Fire Rescue Services (EFRS) to identify hotspot locations for fires in the city of Edmonton during the calendar year 2024. This report is based on the [EFRS Fire Response dataset](https://data.edmonton.ca/d/7hsn-idqi). The information in this dataset is recorded by EFRS staff while co-ordinating the response to each incident. The large volume of incidents and rapid pace of operations in a fire control room mean it is likely the data will contain some errors, but it is not possible to quantify these or assess how they might affect the conclusions drawn in this report. However, it is likely the majority of recording errors are effectively random, and so are unlikely to bias this analysis towards a particular conclusion.


```{r incident-type-count}

# Create a character value containing an list of incidents by type, ordered by
# incident frequency with the most frequent types first
incident_type_count <- incidents |> 
  mutate(
    # Convert categories to lower case
    event_description = str_to_lower(event_description), 
    event_description = case_match(
      event_description, 
      # Collapse all types of fire into one category
      c("outside fire", "fire", "vehicle fire") ~ "fire", 
      # Add label to 'citizen assist' category, which is otherwise hard to 
      # interpret
      "citizen assist" ~ 
        "citizen assist (e.g. animal rescue or person trapped in a lift)",
      .default = event_description
    ),
    # Lump event categories that make up less than 2% of incidents into an
    # 'other' category
    event_description = fct_lump_prop(
      event_description, 
      prop = 0.02, 
      other_level = "other"
    )
  ) |> 
  # Count the number of incidents in each category
  count(event_description, sort = TRUE) |> 
  mutate(
    # Calculate the percentage of incidents in each category
    prop = n / sum(n),
    # Construct a text label
    label = str_glue("{percent(prop, accuracy = 1)} as {event_description}"),
    # Give the first row a slightly different label
    label = if_else(
      row_number() == 1,
      str_replace(label, " as ", " were classified as "), 
      label
    )
  )

daily_incident_count <- incidents |> 
  count(dispatch_date_date) |> 
  pull("n") |> 
  mean()
```


Between 1 January and 31 December 2024, EFRS responded to a total of `r comma(nrow(incidents))` incidents, with an average of `r number(daily_incident_count, accuracy = 1)` per day. Of those, `r knitr::combine_words(pull(incident_type_count, "label"))`. The rest of this report will focus on the `r percent(pluck(filter(incident_type_count, event_description == "fire"), "prop", 1))` of incidents that were categorised as fires.


```{r fire-kde}
#| include: false
#| cache: true

# Create a grid of cells to be used in KDE calculations
kde_grid <- hotspot_grid(neighbourhoods, cell_size = 400, grid_type = "hex")

# Calculate density of fires
fires_kde <- fires |> 
  hotspot_gistar(grid = kde_grid, bandwidth_adjust = 0.3, quiet = TRUE) |> 
  st_intersection(neighbourhoods)

downtown_buffer <- fires_kde |> 
  # Keep only the single grid cell with the highest density of fires
  slice_max(order_by = kde, n = 1, with_ties = FALSE) |> 
  # Find the centre point of that grid cell
  st_centroid() |> 
  # Create a circle 500m around that point
  st_buffer(500)

# Find count of fires within 1km of point with highest density
downtown_count <- hotspot_count(fires, grid = downtown_buffer, quiet = TRUE)
```


```{r fire-map}
#| label: fig-fire-map
#| fig-asp: 0.8
#| fig-cap: "Locations of calls to fires in Edmonton, 2024"
#| out-width: "100%"
#| message: false

fires_kde |>
  filter(gistar > 0, pvalue < 0.05) |> 
  ggplot() +
  # Add base map
  annotation_map_tile(type = "cartolight", zoomin = 1, progress = "none") +
  # Add KDE layer
  geom_sf(aes(fill = kde), colour = NA, alpha = 0.7) +
  # Add district boundaries
  annotation_spatial(data = districts, colour = "grey50", fill = NA) +
  # Add city boundaries
  geom_sf(data = edmonton, colour = "grey25", fill = NA) +
  # Add district labels
  geom_sf_text(
    aes(label = str_wrap(str_to_upper(district), width = 10)), 
    data = districts, 
    colour = "grey50",
    lineheight = 1,
    size = 2.1
  ) +
  # Control appearance of legend fill
  scale_fill_gradient(
    low = rgb(1, 0.8, 0.8), 
    high = rgb(0.6, 0, 0),
    breaks = range(pull(fires_kde, kde)),
    labels = c("lower", "higher")
  ) +
  fixed_plot_aspect(ratio = 1) +
  labs(
    caption = "Data from City of Edmonton and OpenStreetMap",
    fill = "concentration\nof fires"
  ) +
  theme_void()
```


@fig-fire-map shows locations in the city of Edmonton with more fires than would be expected by chance. The map shows that fires are very heavily concentrated in a small area in downtown Edmonton. There were `r comma(pluck(downtown_count, "n"))` fires within 1 kilometre of this point in 2024, around `r percent(pluck(downtown_count, "n") / nrow(fires))` of all fires in Edmonton in that period. This is despite that area being only `r percent(as.numeric(st_area(downtown_buffer) / st_area(st_union(neighbourhoods))), accuracy = 0.1)` of the area of the city.


```{r downtown-kde}
#| include: false
#| cache: true

# Calculate a KDE surface for just the area within 5km of the point with the
# higest density of fires
downtown_kde <- fires |> 
  # `downtown_buffer` stores a buffer 1km around the point of highest density,
  # so we can add 4000m to that to make a 5km buffer
  st_intersection(st_buffer(downtown_buffer, dist = 4000)) |> 
  # Estimate density
  hotspot_gistar(
    cell_size = 100, 
    bandwidth_adjust = 0.3, 
    grid_type = "hex", 
    quiet = TRUE
  )
```


```{r downtown-map}
#| label: fig-downtown-map
#| fig-asp: 0.8
#| fig-cap: "Locations of calls to fires in downtown Edmonton, 2024"
#| out-width: "100%"

ggplot() +
  # Add an invisible layer to the map showing a 3km buffer around the point with
  # the highest density. This layer will be used to control the size of the area
  # shown on the map.
  geom_sf(
    data = st_buffer(downtown_buffer, dist = 2000), 
    colour = NA, 
    fill = NA
  ) +
  # Add base map
  annotation_map_tile(type = "cartolight", zoomin = 1, progress = "none") +
  # Add neighbourhood boundaries
  annotation_spatial(data = neighbourhoods, colour = "grey50", fill = NA) +
  # Add KDE layer, using `annotation_spatial()` so the map doesn't expand to 
  # cover the larger area covered by the KDE layer
  annotation_spatial(
    aes(fill = kde), 
    data = filter(downtown_kde, gistar > 0, pvalue < 0.05), 
    alpha = 0.7, 
    colour = NA
  ) +
  # Add fire station locations
  geom_sf_label(
    aes(label = name), 
    data = filter(fire_stations, name %in% c("01", "02", "05", "21", "22")), 
    colour = "white",
    fill = "red3",
    fontface = "bold",
    label.size = NA,
    size = 2.6
  ) +
  # Add neighbourhood labels
  geom_sf_text(
    aes(label = str_wrap(name, width = 10)),
    data = filter(neighbourhoods, name %in% c("DOWNTOWN", "BOYLE STREET")),
    colour = "grey50",
    lineheight = 1,
    size = 2.1
  ) +
  # Control appearance of legend fill
  scale_fill_gradient(
    low = rgb(1, 0.8, 0.8), 
    high = rgb(0.6, 0, 0),
    breaks = range(pull(fires_kde, kde)),
    labels = c("lower", "higher")
  ) +
  labs(
    subtitle = "Numbered boxes show EFRS stations",
    caption = "Data from City of Edmonton and OpenStreetMap",
    fill = "concentration\nof fires"
  ) +
  theme_void()
```


@fig-downtown-map shows the places within downtown Edmonton with more fires than would be expected by chance. This shows that fires are concentrated around the Downtown and Boyle Street neighbourhoods, as well as along the 107th Avenue corridor.


```{r}
#| label: fig-fire-types
#| fig-asp: 0.5
#| fig-cap: "Frequency of different types of fire in Edmonton, 2024"
#| out-width: "100%"

fires |> 
  # Convert the SF object to a tibble, since we don't need the geometry and it
  # slows down functions lik `count()`
  st_drop_geometry() |> 
  # Count the number of fires of each type
  count(fire_subtype) |> 
  mutate(
    # Order the fire types in order of frequency
    fire_subtype = fct_reorder(fire_subtype, n),
    # Construct a label for each bar
    label = percent(n / sum(n), accuracy = 0.1)
  ) |> 
  # Initiate chart
  ggplot() +
  # Specify how columns in the data should be represented on the chart
  aes(x = n, y = fire_subtype, label = label) +
  # Add labels (behind bars in case labels might cut off the end of the bars)
  geom_label(hjust = 0, label.size = NA) +
  # Add bars
  geom_col(fill = "red3") +
  # Specify appearance of x-axis
  scale_x_continuous(labels = label_comma(), expand = expansion(c(0, 0.1))) +
  # Set chart labels
  labs(
    x = "number of fires",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    # Remove unnecessary gridlines
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```


@fig-fire-types shows the number of fires of each type that were recorded as occurring in Edmonton in 2024. More than two thirds of fires were outdoor fires, e.g. a fire caused by a lightning strike or illegal burning of rubbish. The majority of structure fires were in residential structures.


```{r fire-types-kde}
#| include: false
#| cache: true

fire_type_counts <- fires |> 
  st_drop_geometry() |> 
  count(fire_subtype, sort = TRUE) |> 
  mutate(
    prop = n / sum(n),
    label = str_glue("{fire_subtype} ({percent(prop, accuracy = 0.1)})")
  )

fire_types_kde <- fires |> 
  # Keep just the most-common types of fire
  filter(
    fire_subtype %in% c(
      "outside fire", 
      "residential structure fire", 
      "vegetation/woodland fire", 
      "other structure fire"
    )
  ) |> 
  # Reorder the fire subtypes in descending order of frequency
  mutate(
    fire_subtype = fct_relevel(
      fire_subtype, 
      pull(fire_type_counts, "fire_subtype")
    )
  ) |> 
  group_by(fire_subtype) |> 
  group_modify(
    ~ hotspot_gistar(.x, grid = kde_grid, bandwidth_adjust = 0.3, quiet = TRUE)
  ) |> 
  ungroup() |> 
  st_as_sf() |> 
  st_intersection(edmonton)

fire_types_district_counts <- fires |> 
  # Identify which district each fire occurred in
  st_intersection(districts) |> 
  # Convert to tibble to make `count()` faster
  st_drop_geometry() |> 
  # Count number of fires of each type in each district
  count(fire_subtype, district, sort = TRUE) |> 
  # Calculate proportion of fires of each type occurred in each district
  mutate(prop = n / sum(n), .by = fire_subtype)
```


@fig-types-map shows where the four most-common types of fire concentrated. Outside fires are heavily concentrated downtown, while the other types of fire are more widely distributed. While 
`r percent(pluck(filter(fire_types_district_counts, fire_subtype == "outside fire"), "prop", 1), accuracy = 1)` of all outside fires occurred in the `r pluck(filter(fire_types_district_counts, fire_subtype == "outside fire"), "district", 1)` district (one of `r nrow(districts)` districts in Edmonton), the district with the highest proportion of residential structure fires had only `r percent(pluck(filter(fire_types_district_counts, fire_subtype == "residential structure fire"), "prop", 1), accuracy = 1)` of all such fires in the city. @tbl-district-counts shows the number of fires of each type in each district.


```{r fire-types-map}
#| label: fig-types-map
#| fig-asp: 1.4
#| fig-cap: "Locations of calls to different types of fire in Edmonton, 2024"
#| out-width: "100%"

fire_types_kde |> 
  filter(gistar > 0, pvalue < 0.05) |> 
  left_join(
    fire_type_counts, 
    by = join_by(fire_subtype == fire_subtype)
  ) |> 
  group_by(label) |> 
  group_split() |> 
  map(function (x) {
    
    ggplot(x) +
      # Add base map
      annotation_map_tile(type = "cartolight", zoomin = 1, progress = "none") +
      # Add KDE layer
      geom_sf(aes(fill = kde), colour = NA, alpha = 0.7) +
      # Add district boundaries
      annotation_spatial(data = districts, colour = "grey50", fill = NA) +
      # Add city boundaries
      geom_sf(data = edmonton, colour = "grey25", fill = NA) +
      # Control appearance of legend fill
      scale_fill_gradient(
        low = rgb(1, 0.8, 0.8), 
        high = rgb(0.6, 0, 0),
        breaks = range(pull(x, kde)),
        labels = c("lower", "higher")
      ) +
      labs(
        subtitle = pluck(x, "label", 1),
        fill = "concentration\nof fires"
      ) +
      theme_void() +
      theme(
        panel.border = element_rect(
          colour = "white", 
          fill = NA, 
          linewidth = 1
        )
      )
    
  }) |> 
  wrap_plots(nrow = 2, guides = "collect") + 
  plot_annotation(caption = "Data from City of Edmonton and OpenStreetMap")
```


```{r district-counts-table}
#| label: tbl-district-counts
#| tbl-cap: "Number of fires of each type in each Edmonton district, 2024"
#| warning: false

fire_types_district_counts |> 
  # Count total number of fires of each type (to arrange columns)
  mutate(sum_type = sum(n), .by = fire_subtype) |> 
  mutate(
    # Add space in column names that need it to allow them to be broken over 
    # multiple lines
    fire_subtype = case_match(
      fire_subtype,
      "vegetation/woodland fire" ~ "vegetation/ woodland",
      .default = str_remove(fire_subtype, " fire")
    ),
    # Arrange the fire types in order of frequency
    fire_subtype = fct_reorder(fire_subtype, sum_type)
  ) |>
  # Remove columns we don't want to appear in the table
  select(-prop, -sum_type) |> 
  # Convert to wider format for presentation
  pivot_wider(names_from = fire_subtype, values_from = n, values_fill = 0) |> 
  # Arrange districts in alphabetical order (since we will show fire frequency
  # with colour)
  arrange(district) |> 
  # Convert to gt object
  gt(rowname_col = "district") |> 
  # Add row spanning all columns above header
  tab_spanner(label = "type of fire", columns = everything()) |> 
  # Format numbers
  fmt_number(decimals = 0) |> 
  # Colour cells according to number of fires
  data_color(
    palette = "Reds",
    domain = range(pull(fire_types_district_counts, "n")),
    na_color = "white"
  ) |> 
  # Add a summary row
  grand_summary_rows(
    # Specify summary row should have the label 'Edmonton' with values 
    # calculated as the sum of the number of fires in each column
    fns = "Edmonton" ~ sum(.), 
    # Specify the summary values should be formatted as numbers
    fmt = ~ fmt_number(., decimals = 0)
  ) |> 
  # Add a note at the bottom of the table explaining what the colours mean
  tab_source_note("Column shading shows number of fires")
```

